---
title: "onto2022 for Bioc 2022"
author: "Vincent J. Carey, stvjc at channing.harvard.edu"
date: "`r format(Sys.time(), '%B %d, %Y')`"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{onto2022 for Bioc 2022}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document:
    highlight: pygments
    number_sections: yes
    theme: united
    toc: yes
---

# Introduction: ontologies for information science

- Systematic controlled and versioned vocabularies with fixed-length tags
- Relationships among terms and term referents are formally defined
- Languages or markups for defining and computing with ontologies:
    - RDF/OWL
    - OBO (stanzas)
- Tools for manipulating and using ontologies
    - R ecosystem: ontologyIndex, rols, glitter (RDF querying)
    - Protege
    - ROBOT, pronto

# Two applications

- Precise characterization of phenotypes in GWAS
- Cell type identification with scRNAseq

## gwascat

EBI produces a catalog of GWAS results, retaining those
that are significant and validated.

```{r doont}
suppressPackageStartupMessages({
library(onto2022)
})
if (!exists("newcat")) data(newcat)
newcat
mcols(newcat[1,])$MAPPED_TRAIT_URI
```

EFO is experimental factor ontology.
We have a copy (likely somewhat out-of-date)
in the ontoProc package.

```{r lkefo}
efo = getEFOOnto()
efo$name["EFO:0007789"]
```

## ieugwasr

The Bristol Integrative Epidemiology Unit curates
a large collection of summary statistics for many GWAS.

```{r lkieu, cache=TRUE}
library(ieugwasr)
gi = gwasinfo()
gi
tail(sort(table(gi$ontology)))
```

